Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-06 paper.Rnw
Updates to the method section

\SweaveOpts{echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE}

<<loadpackages, results="hide", echo=FALSE, cache=FALSE>>=
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
@

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable }
\usepackage[square,numbers]{natbib}

%draft mode
\usepackage[colorlinks]{hyperref}
\usepackage[colorinlistoftodos, textwidth=4cm, shadow]{todonotes}
\usepackage[displaymath, tightpage]{preview}

\setlength{\marginparwidth}{1.2in}

%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}

\title{Hippocampal segmentation with MAGeT}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., Voineskos
A., Chakravarty M.M., and the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

The segmentation approach we propose is best understood as an extension of MAGeT
brain adds a preliminary stage to this process in which the template library is
constructed rather than given as input.  As before, MAGeT brain accepts an atlas
library and unlabelled MR subject images as input.  Template library images are
selected from subject images.  The choice of subjects used in the template
library can be made to reflect the neuroanatomy or demographics of the subject
set as a whole (for instance, by sampling equally from cases and controls).

To create the template library, labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images.  As a result, each template library image
has a label from each atlas.  Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure X describes the MAGeT brain
algorithm in pseudocode.

\footnote{Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.}

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{pseudocodesdf}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$subject$}
        \ForAll{$template$}
          \State propagate all labels for template to subject space
          \State store subject labels
        \EndFor
        \State fuse subject labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a subject to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

%  Describe subject datasets in general 
\subsection{Subjects}

\subsubsection{ADNI1 1.5T Screening Dataset}

Clinical, demographic and pre-processed T1-weighted MRI used in this paper were
downloaded by the authors from the ADNI1 database (adni.loni.ucla.edu) between
March 2012 and August 2012. The image dataset download was the "ADNI1:Screening
1.5T" standardized dataset available from ADNI
(http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/)\citep{Wyman2012}.
This image collection contains uniformly preprocessed images which have been
designated to be the "best" after quality control.  All images were acquired
using 1.5T scanners (General Electric Healthcare, Philips Medical Systems or
Siemens Medical Solutions) at multiple sites using the protocol described in
\citep{Jack2008a}.  Representative 1.5T imaging parameters were TR = 2400ms, TI
= 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x 240mm, a
192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel resolution of
1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown in table
\ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

@@

\todo{describe SNT manual segmentations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Check demographics table only includes 1.5T from the Screening}

@@

Before images were registered, the N3 algorithm [16] is first used to minimize
the intensity nonuniformity in each of the atlases and unlabeled subject
images.

@@

We used the following command line when running the ANTS command, 
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual. \todo{include a reference to manual?}

@@

Label fusion is term given to the process of combining the information from
several candidate labellings for an MR image into a single labelling.  In this
paper we explore the benefits of three different fusion methods.

@@

The {\tt xcorr\_vol utility} from the ANIMAL toolkit is used to calculate the 
cross-correlation similarity measure.

@@

This method is similar to cross-correlation weighted voting 
except that image similarity is calculated by the normalised 
mutual information score over the region of interest. 
\todo{is there a reference for NMI voting?}

The {\tt itk\_similarity} utility from the EZMinc toolkit is used to calculate 
the normalised mutual information measure.

%%%%%                        Goodness-of-fit                               %%%%%

\subsection{Goodness-of-fit}

Two segmentation can be compared using the Dice Kappa ($\kappa$) overlap metric:

\begin{equation*} 
\kappa=\frac{2a}{2a+b+c}
\end{equation*}

where $a$ is the number of voxels common to both segmentations and $b+c$ is the
sum of the voxels uniquely identified by either segmentation.  Using this metric
we can compare an automatically generated segmentation to a gold-standard
segmentation.

@@

\todo{In ANIMAL, images aligned to TAL space}

{\bf Label fusion.} Majority vote, Cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote are used.  With the 
weighted majority vote fusion methods, the number of top labels used in the 
fusion was varied from 3 to 20 images.

@@

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}

To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the subject set.

@@

subjects = subset(yr1, VISCODE == 'bl' & Scaled_2 == FALSE)

@@

\bibliographystyle{plain}
\bibliography{references}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
    fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage[round,authoryear]{natbib}

%draft mode
\usepackage[colorlinks]{hyperref}
\usepackage[colorinlistoftodos, textwidth=4cm, shadow]{todonotes}
\usepackage[displaymath, tightpage]{preview}
\setlength{\marginparwidth}{1.2in}

%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

The segmentation approach we propose is best understood as an extension of
multi-atlas segmentation.  MAGeT brain adds a preliminary stage to this process
in which the template library is constructed rather than given as input.  As
before, MAGeT brain accepts an atlas library and unlabelled MR subject images as
input.  Template library images are selected from subject images.  The choice of
subjects used in the template library can be made to reflect the neuroanatomy or
demographics of the subject set as a whole (for instance, by sampling equally
from cases and controls).

To create the template library, labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images. As a result, each template library image
has a label from each atlas. Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure \ref{alg:MAGeT} describes the MAGeT
brain algorithm in pseudocode.  Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$subject$}
        \ForAll{$template$}
          \State propagate all labels for template to subject space
          \State store subject labels
        \EndFor
        \State fuse subject labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a subject to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Subjects}

\subsubsection{ADNI1 1.5T Screening Dataset}
Clinical, demographic and pre-processed T1-weighted MRI used in this paper were
downloaded by the authors from the ADNI1 database (adni.loni.ucla.edu) between
March 2012 and August 2012. The image dataset download was the "ADNI1:Screening
1.5T" standardized dataset available from ADNI
\footnote{
 \url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

For a subset of ADNI1 images, labels of the left and right hippocampi are
available (herein refered to as SNT labels).  Semi-automated hippocampal
volumetry was carried out using a commercially available high dimensional brain
mapping tool (Medtronic Surgical Navigation Technologies, Louisville, CO), that
has previously been validated and compared to manual tracing of the hippocampus
\citep{Hsu2002}. Measurement of hippocampal volume is achieved first by placing
manually 22 control points as local landmarks for the hippocampus on the
individual brain MRI data: one landmark at the hippocampal head, one at the
tail, and four per image (i.e., at the superior, inferior, medial and lateral
boundaries) on five equally spaced images perpendicular to the long axis of the
hippocampus. Second, fluid image transformation is used to match the individual
brains to a template brain \citep{Christensen1997}. The pixels corresponding to
the hippocampus are then labeled and counted to obtain volumes. This method of
hippocampal voluming has a documented reliability of an intraclass coefficient
better than .94 \citep{Hsu2002}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID < 2000,   
               method = "reverse", test=FALSE, overall=TRUE)
latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Before images were registered, the N3 algorithm \citep{Sled1998} is first used
to minimize the intensity nonuniformity in each of the atlases and unlabeled
subject images.

@@

We used the following command line when running the ANTS command, 
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual 
\footnote{\url{https://sourceforge.net/projects/advants/files/Documentation/}}.

@@

Label fusion is a term given to the process of combining the information from
several candidate labellings for an MR image into a single labelling.  In this
paper we explore the benefits of three different fusion methods.

@@

The {\tt xcorr\_vol} utility from the ANIMAL toolkit is used to calculate the
cross-correlation similarity measure.

@@

This method is similar to cross-correlation weighted voting except that image
similarity is calculated by the normalised mutual information score over the
region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility from
the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is used to
calculate the normalised mutual information measure between to images.

%%%%%                        Overlap Measure                               %%%%%

\subsection{Overlap Measure}

The agreement between two segmentations can assessed using the Dice Similarity
Coefficient (DSC):

\begin{equation*} 
DSC=\frac{2|A \cap B|}{|A| + |B|}
\end{equation*}

where $A$ and $B$ are the regions being compared, and the cardinality is the
volume measured in voxels. Using this measure we can compare an automatically
generated segmentation to a gold-standard segmentation.

@@

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote are used.  With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 3 to 20 images.

@@

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
\todo{comparison across field strengths, a la Heckemann2011?}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the subject set.

@@

#TODO: include Scaled_2? 
subjects = subset(yr1, VISCODE == 'bl' & Scaled_2 == FALSE)

@@

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}
